---
title: "top10_final"
author: "Joseph Redinger"
date: "8/30/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **Script Summary:** Fits and analyzes best GLMs for counts of Impact Factors calculated from GLANSIS organisms risk assessments. Hurdle model was initially considered best fit.  

### **Paper:** The Great Lakes’ Most Unwanted: Characterizing the Impacts of the Top Great Lakes Aquatic Invasive Species - El Lower, Rochelle Sturtevant, Susannah Iott,Felix Martinez, Ed Rutherford, Doran M. Mason and Ashley K. Elgin  

#### **Libraries**
```{r, message = FALSE, warning = FALSE}
library(knitr)
library(dplyr)
library(ggplot2) # plot figures
library(MASS) # run glm with negative binomial distribution
library(pscl) # run zero-inflation model
library(boot) # bootstrapping confidence interval
library(countreg) # plot model dispersion
library(ggpubr) # arrange ggplots for figure
library(ggpattern) # add pattern to black and white pie charts
```


#### **Upload data**

##### This analysis looked only at the species nonindigenous to the Great Lakes basin which scored as ‘invasive’ for either environmental or socioeconomic impact using the original OIA process.  All species with IF of 0 & 1 were removed (impact factors range from 2-72).This effectively excludes all species classified as ‘no significant impact’ and ‘unknown’.


```{r}
dat = read.csv("glansis_species_impact_scores.csv", header = T)
dat = dat[which(dat['status'] == 'established'),]
dat = dat[which(dat['IF.score'] > 1),]

species_remove = c('Aeromonas salmonicida', 'Cyclops strenuus', 'Ictiobus niger', 'Lupinus polyphyllus', 'Nasturtium officinale', 'Pylodictis olivaris', 'Radix auricularia', 'Rorippa sylvestris', 'Rumex longifolius', 'Schyzocotyle acheilognathi', 'Ulva (Enteromorpha) prolifera')

dat <- dat[!(dat$scientific.name %in% species_remove), ]
dat
```

```{r}
sequence = paste(seq(2, 54, by = 2), "-", seq(3, 55, by = 2))

breaks = seq(2, 56, by = 2)
intervals = cut(dat$IF.score, breaks = breaks, include.lowest = TRUE, right = FALSE)
frequency_table = table(intervals)

color_labels = c(rep("black", 2), rep("grey", 7), rep("white", 18))
#color_labels = c(rep("white", 2), rep("grey", 5), rep("black", 20))

barplot_df = data.frame(categories = sequence, frequency = as.vector(frequency_table), color = color_labels)
print(barplot_df)

# Create order that is used in bar plot
barplot_df$categories = factor(barplot_df$categories, levels = barplot_df$categories)

# Bar plot
inv_barplot = ggplot(barplot_df, aes(x = categories, y = frequency, fill = color)) +
  geom_hline(yintercept = seq(0, 25, by = 1), color = "gray88", linetype = "solid") +
  geom_hline(yintercept = seq(0, 25, by = 5), color = "gray40", linetype = "solid") +
  geom_bar(stat = "identity", color = "black", width = 0.8, size = 0.75) +
  xlab( 'Impact Factor') +
  ylab('Number of Species') +
  scale_fill_identity() +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_line(color = "gray", linetype = "solid"),
        axis.text.x = element_text(size = 10, color = "black", face = "bold", angle = 270, hjust = -0.01, vjust = 0.3, margin = margin(t = -10)),
        axis.text.y = element_text(size = 12, color = "black", face = "bold"),
        axis.title.x = element_text(size = 16, face = "bold", margin = margin(t = 1, unit = "lines")),
        axis.title.y = element_text(size = 16, face = "bold", margin = margin(r = 1, unit = "lines"))
  )
inv_barplot

ggsave(inv_barplot, file = "impact.factor.barplot.jpg", width = 20, height = 12, units = c("cm"), dpi = 400)
#ggsave(inv_barplot, file = "impact.factor.barplot.pdf", width = 20, height = 12, units = c("cm"), dpi = 400)
```

### Data Exploration
``` {r, include = FALSE}
# top10
top10 = dat[1:10,]


head(top10, 5)
names(top10)
dim(top10)
str(top10)
```

```{r, include = FALSE}
# all nonindigenous species
all = dat[11:nrow(dat),]
all

head(all, 5)
names(all)
dim(all)
str(all)
```

### Create tables - group by origin & taxa & vector
```{r, include = FALSE}
# Nonindigenous species
all.org = as.data.frame(table(all$origin))
#all.org = as.data.frame(table(all$continent.pub))

all.org = all.org %>% 
      rename("origin" = "Var1") %>%
      rename("freq.all"="Freq")

# Top 10 species
top10.org = as.data.frame(table(top10$origin))

top10.org = top10.org %>% 
      rename("origin" = "Var1") %>%
      rename("freq.top10" = "Freq")

# Combine tables
df.org = full_join(all.org, top10.org, by = join_by("origin" == "origin")) %>% 
      replace(is.na(.), 0)

```

```{r, include = FALSE}
# Nonindigenous species
all.taxa= as.data.frame(table(all$taxa))
#all.taxa = as.data.frame(table(all$taxon.pub))

all.taxa = all.taxa %>% 
      rename("taxa" = "Var1") %>%
      rename("freq.all"="Freq") 

# Top 10 species
top10.taxa = as.data.frame(table(top10$taxa))
#top10.taxa = as.data.frame(table(top10$taxon))

top10.taxa = top10.taxa %>% 
       rename("taxa" = "Var1") %>%
       rename("freq.top10"="Freq")

# Combine tables
df.taxa = full_join(all.taxa, top10.taxa, by = join_by("taxa" == "taxa")) %>% 
      replace(is.na(.), 0)

```

```{r, include = FALSE}
# Nonindigenous species
all.vector = as.data.frame(table(all$vector))

all.vector = all.vector %>% 
      rename("vector" = "Var1") %>%
      rename("freq.all"="Freq") 

# Top 10 species
top10.vector = as.data.frame(table(top10$vector))

top10.vector = top10.vector %>% 
       rename("vector" = "Var1") %>%
       rename("freq.top10"="Freq")

# Combine tables
df.vector = full_join(all.vector, top10.vector, by = join_by("vector" == "vector")) %>% 
      replace(is.na(.), 0)

```

#### Continent of origin data
```{r}
df.org
df.taxa
df.vector
```

### Chi-square test of independence 
```{r, warning = FALSE}
# origin chi-square test
print(chisq.test(df.org[2:3]))

# taxa chi-square test
print(chisq.test(df.taxa[2:3]))

# taxa chi-square test
print(chisq.test(df.vector[2:3]))

```